SU($n$) ferromagnets near the Heisenberg critical point

### 2020-04-04 collective_sun/sun_perturbation.tex
re-wrote the introduction to the Heisenberg model

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multilevel generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by larger, non-perturbative modifications to the
SU($n$) Heisenberg model.  While the latter simulations generally
accumulate uncontrolled truncation errors, we identify criteria under
which numerical results correctly capture the qualitative dynamical
behavior of a system.

@@

% todo: give example of SU(2)

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu\in\ZZ_n} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\C_N\p{2}$ is the set of all
subsets (``choices'') of two spins out of $N$; $h_{pq}$ are scalar
coupling constants; $\mu,\nu$ index states in an orthonormal basis for
the $n$-dimensional Hilbert space of a single spin; $\ZZ_n$ is the set
of integers modulo $n$; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\nu$ from $\ket\mu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is
essentially a multilevel generalization of the SU(2)-symmetric spin
interaction $\v s_p\c\v s_q$, with $\v s\equiv\p{\v s_\x,s_\y,s_\z}$,
$s_\alpha\equiv\sigma_\alpha/2$, and $\sigma_\alpha$ a single-spin
Pauli operator.  This permutation operator is equivalently the
multilevel SWAP operation, which merely swaps the state of two spins.

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of permutationally symmetric (PS) states
that are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{pq}$.  We will assume that all $h_{pq}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is permutationally symmetric
(e.g.~a spin-polarized state).  We will also assume that the PS
manifold $\M_0$ is gapped away from all orthogonal states by an
interaction energy $\Delta_{\t{gap}}>0$.  Power-law couplings of the
form $h_{pq}\sim-1/\abs{p-q}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap $\Delta_{\t{gap}}$
for long-range interactions with $\alpha\le D$ (see Appendix
\ref{sec:spectral_gap}).

The PS manifold $\M_0$ is spanned by a basis of states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu$ of the single-spin state $\mu$, and $\sum_\mu m_\mu=N$.  We
denote the set of all such assignments of $N$ (identical) spins to $n$
(distinct) states by $\A_n\p{N}$, such that our standard basis for
$\M_0$ is $\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$ is
then the number of such assignments, which is straightforward to
determine with a standard ``stars and bars'' calculation:
$\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.  In the case of
$n=2$, for example, the PS manifold $\M_0$ is precisely the Dicke
manifold of $N$ qubits, spanned by ${N+2-1\choose2-1}=N+1$ Dicke
states $\ket{m_0,m_1}$ for which $m_0+m_1=N$ and the integer $m_0$
($m_1$) indicates the total occupation number of qubit state $\ket{0}$
($\ket{1}$).  The energy of any PS state $\ket\psi\in\M_0$ with
respect to the interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\set{p,q}\in\C_N\p{2}} h_{pq}.
\end{align}

@@

Here we show that power-law SU($n$)-symmetric interactions of the form
in \eqref{eq:H_0} with $h_{pq}=-h/\abs{p-q}^\alpha$ yield a
non-vanishing many-body interaction energy gap when $\alpha\le D$,
where $D$ is the dimension of the lattice.  On a periodic lattice of
$N=L^D$ spins, the spectral gap of the interaction Hamiltonian is (see
Appendix \ref{sec:single_body_eigenstates})
\begin{align}
  \Delta
  = \sum_{d\in\ZZ_L^D} h_{0,d} \sp{\cos\p{d \c k_{\t{SE}}}-1}
  = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d \c k_{\t{SE}}}}{\abs{d}^\alpha},
\end{align}
where $\ZZ_L$ denotes the integers modulo $L$;
$k_{\t{SE}}\in\ZZ_L^D\times2\pi/L$ is a wavenumber for a
singly-excited spin-wave state; and the norm $\abs{d}$ for a vector
$d$ on a periodic lattice is implicitly understood to mean the
smallest Euclidean distance of $d$ from a fixed origin.  In order to
yield the smallest excitation energy $\Delta$, the wavenumber
$k_{\t{SE}}$ should maximize the contribution of the cosine term
above, which is achieved by a wavenumber that minimizes the
oscillations of this term when integrated over the entire lattice.  A
suitable candidate for a minimal wavenumber is
$k_{\t{SE}}=\p{2\pi/L,0,0,\cdots}$, which corresponds to an excitation
energy
\begin{align}
  \Delta = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d_12\pi/L}}{\abs{d}^\alpha}.
\end{align}
Defining $\epsilon\equiv2/L$ and a rescaled domain symmetrized about
$0$, $\SS_\epsilon\simeq\ZZ_L/\epsilon$ with
$\SS_\epsilon\subset\sp{-1,1}$, we substitute $x\simeq\epsilon d$ to
get
\begin{align}
  \Delta
  = h \sum_{\substack{x\in\SS_\epsilon^D\\\abs{x}\ge\epsilon}}
  \f{1-\cos\p{\pi x_1}}{\abs{x/\epsilon}^\alpha}
  = h \epsilon^{\alpha-D} \sum_{\substack{x\in\SS_L^D\\\abs{x}\ge\epsilon}}
  \epsilon^D \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha}.
\end{align}
As $\epsilon\to0$ the discrete sum over $x$ is well approximated by an
integral that avoids an infinitesimal region about the origin, i.e.
\begin{align}
  \Delta = h \epsilon^{\alpha-D} \I_{D\epsilon},
  &&
  \I_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha},
\end{align}
where the integral $\I_{D\epsilon}$ is defined using the interval
$\TT_a\equiv\p{-a,a}$.  The integrand of $\I_{D\epsilon}$ is strictly
positive and well-behaved on the entirety of its domain except for the
origin, where depending on the value of $\alpha$ the integrand may
vanish or diverge as $\abs{x}\to0$.  Together, these facts mean that
\begin{align}
  \I_{D\epsilon} \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \Delta \stackrel{\epsilon\to0}{\sim} h \epsilon^{\alpha-D-\gamma},
\end{align}
for some $\gamma\ge0$, which implies that gap $\Delta$ is
non-vanishing when $\alpha\le D\le D+\gamma$.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\G}{\mathcal{G}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\T}{\mathcal{T}}

@@

In this work, we generalize some of the techniques used to harness
SU(2)-symmetric spin interactions for metrological advantage to the
case of SU($n$).  In doing so, we develop a low-energy effective
theory of gapped SU($n$) ferromagnets with weakly broken SU($n$)
symmetry, e.g.~by the presence of external fields or additional
interaction terms.  The restriction of our theory to ferromagnetic
interactions can be relaxed to the assumption of
\begin{enumerate*}
\item an initial state in the manifold of permutationally-symmetric
  states, a multi-level generalization of the Dicke manifold for
  two-level spins, together with the assumption that
\item this manifold is gapped away from states that break
  permutational symmetry.
\end{enumerate*}
In addition to working out the effective Hamiltonian induced by weak
SU($n$) symmetry breaking through second order in perturbation theory,
we develop a diagrammatic method to systematically compute all matrix
elements of arbitrary spin operators restricted manifolds of weakly
broken permutation symmetry, thereby enabling the numerical simulation
of dynamics induced by larger, non-perturbative modifications to the
SU($n$) Heisenberg model.  While the latter simulations generally
accumulate uncontrolled truncation errors, we identify criteria under
which numerical results correctly capture the qualitative dynamical
behavior of a system.

@@

Before getting to the full SU($n$) Heisenberg model, we first revisit
the more common case of SU(2).  The SU(2) Heisenberg model consists of
$N$ two-level spins on a lattice, with spin-aligning interactions
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}}
  h_{ij} \m s_i \c\m s_j,
  \label{eq:H_SU2}
\end{align}
where
$\m s_i\equiv\p{\sigma_\x^{(i)},\sigma_\y^{(i)},\sigma_\z^{(i)}}/2$ is
a spin vector defined with Pauli operators $\sigma_\alpha^{(i)}$
acting on the spin at lattice site $i$.  In the traditional language
of spin physics, each term of this Hamiltonian essentially imposes an
energy penalty of $h_{ij}$ to the formation of a spin 0 singlet state
$\ket{\up\dn}-\ket{\dn\up}$ between spins $i$ and $j$, favoring spin
triplet states with total spin 1.  Another interpretation of the
Hamiltonian in \eqref{eq:H_SU2} is illuminated by the observation that
the two-spin operator $\m s_i\c\m s_j$ is, up to an overall constant,
essentially the two-qubit SWAP operation $\Pi_{ij}$.  We will also
refer to $\Pi_{ij}$ as {\it permutation operator}, because its action
on the Hilbert space $\H\equiv\bigotimes_{k=1}^N\H_k$ of $N$ spins
with individual Hilbert spaces $\H_k$ is to permute the tensor factors
of $\H$ associated with spins $i$ and $j$.  Up to identity terms with
no physical consequence, we then have
\begin{align}
  H_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij} \Pi_{ij},
  &&
  \Pi_{ij} \equiv \sum_{\substack{\t{spin states}\\\mu,\nu}}
  s_{\mu\nu}^{(i)} s_{\nu\mu}^{(j)}
  = \sum_{\substack{\t{spin states}\\\mu,\nu}} \op{\mu\nu}{\nu\mu}_{ij},
  \label{eq:H_0}
\end{align}
% todo: insert visualization of SWAP operation
where $s_{\mu\nu}^{(i)}\equiv\op{\mu}{\nu}_i$ is a spin-transition
operator for spin $i$, and $\mu,\nu\in\set{\up,\dn}$ index the states
of a single spin.  In addition to providing a nice interpretation of
the Hamiltonian $H_0$ as essentially swapping the states of spins $i$
and $j$ at a rate $h_{ij}$, the definition of the SU(2)-symmetric
Heisenberg model in in \eqref{eq:H_0} is straightforward to generalize
to SU($n$) with arbitrary $n$.  That is, rather than having
$\mu,\nu\in\set{\up,\dn}$, we simply let
$\mu,\nu\in\set{1,2,\cdots,n}$ index orthogonal states (sometimes
called {\it colors} or {\it flavors}) of an $n$-level spin.

If the coefficients $h_{ij}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of $H_0$ consists
of spin-polarized, {\it permutationally symmetric} (PS) states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{ij}$.  We will assume that all $h_{ij}<0$ throughout these
notes, but note that this restriction can be relaxed to the assumption
that the initial state of all spins is a permutationally-symmetric
state.  We will also assume that the PS manifold $\M_0$ is gapped away
from all orthogonal states by an interaction energy
$\Delta_{\t{gap}}>0$.  Power-law couplings of the form
$h_{ij}\sim-1/\abs{\m r_i-\m r_j}^\alpha$ on a $D$-dimensional
lattice, for example, always yield a non-vanishing spectral gap
$\Delta_{\t{gap}}$ for long-range interactions with $\alpha\le D$ (see
Appendix \ref{sec:spectral_gap}).  Here $\m r_i$ denotes the position
of lattice site $i$.

In the case of SU(2), the PS manifold $\M_0$ is precisely the {\it
  Dicke manifold} spanned by the PS states $\ket{m_\up,m_\dn}$ with a
definite (integer) number $m_\mu\ge0$ of spins occupying the
single-spin state $\mu\in\set{\up,\dn}$.  If there are $N$ spins
total, then Dicke states $\ket{m_\up,m_\dn}$ must have
$m_\up+m_\dn=N$.  In the case of multi-level spins, the PS manifold
$\M_0$ is similarly spanned by states
$\ket{m}=\ket{m_1,m_2,\cdots,m_n}$ with a definite occupation number
$m_\mu\ge0$ of the single-spin state $\mu$, with $\sum_\mu m_\mu=N$.
These states are essentially labeled by an assignment
$m=\p{m_1,m_2,\cdots,m_n}$ of $N$ (identical) spins to $n$ (distinct)
states.  We denote the set of all such assignments by $\A_n\p{N}$,
such that our standard basis for $\M_0$ is
$\set{\ket{m}:m\in\A_n\p{N}}$.  The dimension of $\M_0$, or
equivalently the ground-state degeneracy of $H_0$, is straightforward
to determine with a standard ``stars and bars'' calculation:
$\dim\M_0=\abs{\A_n\p{N}} = {N+n-1 \choose n-1} \sim N^{n-1}$.  The
energy of any PS state $\ket\psi\in\M_0$ with respect to the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0} is
\begin{align}
  E_0 = \sum_{\substack{\t{spin pairs}\\\set{i,j}}} h_{ij}
  = \sum_{i<j} h_{ij},
\end{align}
where $i,j\in\set{1,2,\cdots,N}$ index individual lattice sites.

@@

Here we show that power-law SU($n$)-symmetric interactions of the form
in \eqref{eq:H_0} with $h_{pq}=-h/\abs{p-q}^\alpha$ yield a
non-vanishing many-body interaction energy gap when $\alpha\le D$,
where $D$ is the dimension of the lattice.  On a periodic lattice of
$N=L^D$ spins, the spectral gap of the interaction Hamiltonian is (see
Appendix \ref{sec:single_body_eigenstates})
\begin{align}
  \Delta
  = \sum_{d\in\ZZ_L^D} h_{0,d} \sp{\cos\p{d \c k_{\t{SE}}}-1}
  = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d \c k_{\t{SE}}}}{\abs{d}^\alpha},
\end{align}
where $\ZZ_L$ denotes the integers modulo $L$;
$k_{\t{SE}}\in\ZZ_L^D\times2\pi/L$ is a wavenumber for a
singly-excited spin-wave state; and the norm $\abs{d}$ for a vector
$d$ on a periodic lattice is implicitly understood to mean the
smallest Euclidean distance of $d$ from a fixed origin.  In order to
yield the smallest excitation energy $\Delta$, the wavenumber
$k_{\t{SE}}$ should maximize the contribution of the cosine term
above, which is achieved by a wavenumber that minimizes the
oscillations of this term when integrated over the entire lattice.  A
suitable candidate for a minimal wavenumber is
$k_{\t{SE}}=\p{2\pi/L,0,0,\cdots}$, which corresponds to an excitation
energy
\begin{align}
  \Delta = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d_12\pi/L}}{\abs{d}^\alpha}.
\end{align}
Defining $\epsilon\equiv2/L$ and a rescaled domain symmetrized about
$0$, $\SS_\epsilon\simeq\ZZ_L/\epsilon$ with
$\SS_\epsilon\subset\sp{-1,1}$, we substitute $x\simeq\epsilon d$ to
get
% todo: define \simeq
\begin{align}
  \Delta
  = h \sum_{\substack{x\in\SS_\epsilon^D\\\abs{x}\ge\epsilon}}
  \f{1-\cos\p{\pi x_1}}{\abs{x/\epsilon}^\alpha}
  = h \epsilon^{\alpha-D} \sum_{\substack{x\in\SS_L^D\\\abs{x}\ge\epsilon}}
  \epsilon^D \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha}.
\end{align}
As $\epsilon\to0$ the discrete sum over $x$ is well approximated by an
integral that avoids an infinitesimal region about the origin, i.e.
\begin{align}
  \Delta = h \epsilon^{\alpha-D} \I_{D\epsilon},
  &&
  \I_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha},
\end{align}
where the integral $\I_{D\epsilon}$ is defined using the interval
$\TT_a\equiv\p{-a,a}$.  The integrand of $\I_{D\epsilon}$ is strictly
positive and well-behaved on the entirety of its domain except for the
origin, where depending on the value of $\alpha$ the integrand may
vanish or diverge as $\abs{x}\to0$.  Together, these facts mean that
\begin{align}
  \I_{D\epsilon} \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \Delta \stackrel{\epsilon\to0}{\sim} h \epsilon^{\alpha-D-\gamma},
\end{align}
for some $\gamma\ge0$, which implies that gap $\Delta$ is
non-vanishing when $\alpha\le D\le D+\gamma$.
